---
title: "Spatial networks - First attempt"
author: "Michael Dorman"
date: "March 12, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document describes a preliminary workflow for familiarizing with using `igraph` to represent a spatial network. `igraph` is a package for network analysis, with the main data structure (`igraph`) used to represent a network. The term *spatial network* hereby refers to any network where both the **vertices** and the **edges** are associated with spatial features. For example, in a transportation network the **vertices** may correspond to junctions, bus stops, etc., while **edges** may correspond to the road segments connecting them. 

In the following simple network, vertices correspond to US state centroids, while edges are straight line connections between the centroids. Not all possible vertex combinations are connected, however. We are going to connect only those centroids that correspond to states sharing a common boundary.

As a first step, let us obtain a `SpatialPolygonsDataFrame` object of US states -

```{r, message=FALSE}
library(rgdal)
library(rgeos)
library(igraph)

# Read states layer from a GeoJSON file
url = "https://raw.githubusercontent.com/alignedleft/d3-book/master/chapter_12/us-states.json"
state = readOGR(url, "OGRGeoJSON", stringsAsFactors = FALSE, verbose = FALSE)

# Transform to US-Atlas projection
state = spTransform(state, "+proj=laea +lat_0=45 +lon_0=-100 +x_0=0 +y_0=0 +a=6370997 +b=6370997 +units=m +no_defs")

# Assign state names to 'row.names'
row.names(state) = state$name
```

The attribute table of the `state` layer has the following structure. Note that we assigned state names to the `row.names` property. These are going to be the vertex names in the network.

```{r}
head(state@data)
```

And here is a plot of the `state` layer:

```{r}
plot(state)
```

One method to create an `igraph` object is to convert an adjacency matrix into `igraph` using function `graph_from_adjacency_matrix`. Since we would like to create edges only between those states that share a common boundary, our (52×52) matrix will have the value of `0` for all state×state combinations where there is no common boundary. Moreover, for those state pairs that *do* share a common boundary, we will give each edge a weight equal to the straight line distance between the centroids. That way, our network is going to be *weighted*, with weight representing an approximate distance among locations within the given states. 

```{r}
# Calculate 'SpatialPoints' layer of state centroids
ctr = gCentroid(state, byid = TRUE)

# Calculate pairwise logical matrix on feature intersection
m = gIntersects(state, byid = TRUE)

# Calculate pairwise distance matrix (in km)
d = gDistance(ctr, byid = TRUE) / 1000

# Assign 0 for 'no intersection', otherwise keep centroid distance
d[!m] = 0
```

Here are the first 5 rows and columns of the resulting matrix. Zero values mean that the states **do not** share a common border; non-zero values represent geographical distance (in km) between state centroids. For example, we can see that among the first five states only California and Arizona share a common boundary, and the distance between their centroids is `r round(d["California", "Arizona"])` km. 

```{r}
d[1:5, 1:5]
```

Creating the `igraph` object goes as follows:

```{r}
g = graph_from_adjacency_matrix(d, mode = "undirected", weighted = TRUE)
```

Thanks to the `mode = "undirected"` argument, the geographical distance is now stored in the `weight` property of each edge:

```{r}
E(g)$weight[1:5]
```

Additionally, we can keep the `x` and `y` coordinates as **vertex** properties:

```{r}
V(g)$x = coordinates(ctr)[, 1]
V(g)$y = coordinates(ctr)[, 2]
```

The network can be visualized with `plot.igraph`. Since vertices have `x` and `y` properties (which are spatial coordinates), their placement corresponds to spatial location: 

```{r}
plot(g)
```

This representation does not cover all spatial uses, however. We would also like to have the vertices and edges associated with features (points and lines, in this example). For example, when we employ network analysis methods such as **shortest path** or **community detection** we would like to get the corresponding spatial subsets, such as a line representing the path in space, or the a subset of vertices representing a community. How can we do this? A basic approach is outlined below.

To represent the vertices we can simply use the centroid layer `ctr` created earlier. (only the contigous USA states are plotted below for clarity)

```{r}
main = state[!state$name %in% c("Alaska", "Hawaii", "Puerto Rico"), ]
plot(main, border = "grey")
plot(ctr, add = TRUE, pch = 1)
```

Creating the edge representation is a little more complicated. To do that, we need to make a line layer where each feature is a line (composed of two points) corresponding to a given vertex in our `igraph` object `g`. 

The first step is to summarize edge name pairs for each vertex in `g`:

```{r}
edges = get.edgelist(g)
edges = data.frame(
  v1 = edges[,1], 
  v2 = edges[,2], 
  stringsAsFactors = FALSE
  )
head(edges)
```

For example, we see that the first two edges connect Alabama with Florida and with Georgia, respectively:

```{r}
E(g)[1:2]
```

The second step is to make a `SpatialLines` object representing those edges. This can be done by sequentially connecting point-pairs from `ctr` with a `for` loop going over the rows in the `edges` table. Line construction itself is hereby performed with the `shadow::ray` function:

```{r}
# Start with empty list
l = NULL 

for(i in 1:nrow(edges)) # For each edge
  l = c(
    l, 
    # Create 'SpatialLines' between the corresponding vertices
    shadow::ray( 
      ctr[edges$v1[i], ], 
      ctr[edges$v2[i], ]
      )
    )

# Combine all lines into one 'SpatialLines' object
l = do.call(rbind, l) 
```

Here is the result:

```{r}
plot(main, border = "grey")
plot(ctr, add = TRUE, pch = 1)
plot(l, add = TRUE)
```

At this point we have:

* A network object `g` representing the relation between vertices (states).
* A `SpatialPoints` object `ctr` representing vertex locations (state centroids).
* A `SpatialLines` object `l` representing edge locations (straight lines between state centroids).

For now, the association between these three objects is based on the following:

* The order of `ctr` corresponds to the order of vertices in `g`.
* The order of `l` corresponds to the order of edges in `g`.

Thus, any statistical or descriptive calculation involving vertices and edges from `g` can later on be expressed using their spatial representation.

For example, we can find out the shortest path from Texas to Washington, in terms of vertex and edge IDs:

```{r}
# Find shortest path
p = shortest_paths(g, "Texas", "Washington", output = "both")

# The 'vpath' component contains the vertex IDs on the path
p$vpath
unlist(p$vpath)

# The 'epath' component contains the edge IDs on the path
p$epath
unlist(p$epath)
```

The corresponding point and line layer can be created by subsetting the full layers:

```{r}
p_vertices = unlist(p$vpath)
p_edges = unlist(p$epath)
p_ctr = ctr[p_vertices, ]
p_l = l[p_edges, ]
```

Let's visualize the path's vertices and edges in red, with the full network in the background:

```{r}
plot(main, border = "grey")
plot(p_l, add = TRUE, col = "red", lwd = 3)
plot(p_ctr, add = TRUE, col = "red", pch = 16, cex = 1)
plot(l, add = TRUE)
plot(ctr, add = TRUE, pch = 1)
```

We can see that the first step in the path goes from Texas to New Mexico. What if these two states suddenly became 'more distant' from each other?

```{r}
E(g)[from("Texas") & to("New Mexico")]$weight = 2000
```

The new path would then bypass New Mexico, going through Oklahoma instead:

```{r}
p = shortest_paths(g, "Texas", "Washington", output = "both")
p$vpath
p$epath
p_vertices = p$vpath[[1]]
p_edges = p$epath[[1]]
```

And here is the new plot. The vertex we made more 'distant' is marked in yellow:

```{r}
p_ctr = ctr[p_vertices, ]
p_l = l[p_edges, ]

plot(main, border = "grey")
plot(p_l, add = TRUE, col = "red", lwd = 3)
plot(l[E(g)[from("Texas") & to("New Mexico")], ], add = TRUE, col = "yellow", lwd = 3)
plot(p_ctr, add = TRUE, col = "red", pch = 16, cex = 1)
plot(l, add = TRUE)
plot(ctr, add = TRUE, pch = 1)
```

That's it for now, thanks for reading!




